### 2025-02-24 nondirectional_ms_2024.Rmd
Minor changes

abstract: |
  A wave of criticisms and replication failures is currently challenging claims about the scope of unconscious perception and cognition. Such failures to find unconscious processing effects at the population level may reflect the absence of individual-level effects, or alternatively, the averaging out of individual-level effects with opposing signs. Importantly, only the first suggests that consciousness may be necessary for the tested process to take place. To arbitrate between these two possibilities, we tested previously collected data where unconscious processing effects were not found (26 effects from 470 participants), using five frequentist and Bayesian tests that are robust to individual differences in effect signs. By and large, we found no reliable evidence for unconscious effects being masked by individual differences. In contrast, when we examined 135 non-significant effects from other domains, two novel non-parametric tests did reveal effects that were hidden by opposing individual results, though as we show, some of them might be driven by design-related factors. Taken together, five analysis approaches provide strong evidence for the restricted nature of unconscious processing effects not only across participants, but also across different trials within individuals. We provide analysis code and best-practice recommendations for testing for non-directional effects.

@@

Second, the *qualitative individual differences* approach (@rouder2021a; @haaf2019a; henceforth QUID) quantifies the relative support for the presence of “qualitative differences” in effects, that is, inter-individual differences in effect signs, by performing a Bayesian model comparison over a family of hierarchical models with different constraints [@haaf2019a]. Specifically, QUID models reaction times (RTs) using a standard linear model, where individual-level effects are treated as random effects. Then, to quantify the support for an effect while allowing for qualitative individual differences, it relies on Bayesian Factors (BFs). The BF analysis compares evidence of a model that poses no constraints over the magnitude of individual-level effects, with a model in which all participants show no effect (see @rouder2021a for further model comparisons, aimed at examining different hypotheses regarding the character of population variability in effects).

@@

We applied the tests to our simulated data. For QUID, we used the default priors from the original publication [@rouder2021a]. For GNT and OANOVA, we used an $\alpha$ of `r oanova_alpha` to examine individual-level and group-level effects. For QUID, we considered $BF>`r bf_criteria`$ as evidence for an effect, $BF< \frac{1}{`r bf_criteria`}$ as evidence for no effect (*global null*), and values between these thresholds ($\frac{1}{`r bf_criteria`}\leq BF\leq `r bf_criteria`$) as inconclusive [@jeffreys1998a]. Reassuringly, all tests were able to differentiate between the two simulated scenarios, providing very strong evidence for an effect in the *non-directional differences* scenario, but not in the *global null* one. Specifically, the prevalence of effects on RT was clearly above the expected proportion of significant effects ($\alpha_{individual}$)  in the *non-directional differences* simulation (using a two-sided t-test for the individual-level test; `r round(100 * nde_gnt_res$stat)`% significant effects, one-sided $CI_{95}$ = [`r round(100 * nde_gnt_res$ci_low)`, `r round(100 * nde_gnt_res$ci_high)`], p `r apa_p(nde_gnt_res$p)`), but this proportion was not higher than $\alpha_{individual}$ in the *global null* simulation (`r round(100 * sn_gnt_res$stat)`% significant effects, one-sided $CI_{95}$ = [`r round(100 * sn_gnt_res$ci_low)`, `r round(100 * sn_gnt_res$ci_high)`], p =`r apa_p(sn_gnt_res$p)`). Using the QUID method, a random effects model with individual-level effects was overwhelmingly preferred in the *non-directional differences* simulation ($BF$= `r formatC(nde_bf, format = 'e', 2)`), but a null model was preferred in the *global null* simulation ($BF$= `r round(sn_bf,2)`). Similarly, the OANOVA test revealed significant results in the *non-directional differences* scenario (F(`r nde_OANOVA_res$df_num`, `r nde_OANOVA_res$df_denom`) = `r round(nde_OANOVA_res$F,2)`, p `r paste0(ifelse(nde_OANOVA_res$p >= .001, '=', ''), apa_p(nde_OANOVA_res$p))`), and a non-significant effect in the *global null* simulation (F(`r sn_OANOVA_res$df_num`, `r sn_OANOVA_res$df_denom`) = `r round(sn_OANOVA_res$F,2)`, p `r paste0(ifelse(sn_OANOVA_res$p >= .001, '=', ''),apa_p(sn_OANOVA_res$p))`).

@@

Yet, the reviewed approaches also have some limitations that make it harder to draw firm conclusions based on their results. First, in contrast to frequentist tests within the Null Hypothesis Statistical Testing (NHST) tradition, QUID is designed to quantify relative evidence, and as such provides only weak^[Specifically, p($BF_{10}>x|H_0$) is never higher than $\frac{1}{x})$. Proof: assume that $BF_{10} > 3$ for observed values $v \in V$. That is, for every $v \in V$, $p(v|H_1)>3p(v|H_0)$. The probability of making a type-1 error $\alpha$ equals $p(v \in V|H_0)$. It follows that $p(v \in V|H_1)>3\alpha$. Since this probability cannot be higher than 1, $\alpha$ cannot exceed $\frac{1}{3}$.] control over long-term error rates (the probability of finding a false positive result or missing a true result over an infinite number of tests, with the former being more critical to our point here). Such error control promises a much-needed ‘fool-proof’ method to infer the existence of unconscious processing effects without making too many mistakes in the long run [@lakens2020a; @kelter2021a].

>>>>

abstract: |
  A wave of criticisms and replication failures is currently challenging claims about the scope of unconscious perception and cognition. Such failures to find unconscious processing effects at the population level may reflect the absence of individual-level effects, or alternatively, the averaging out of individual-level effects with opposing signs. Importantly, only the first suggests that consciousness may be necessary for the tested process to take place. To arbitrate between these two possibilities, we tested previously collected data where unconscious processing effects were not found (26 effects from 470 participants), using five frequentist and Bayesian tests that are robust to individual differences in effect signs. By and large, we found no reliable evidence for unconscious effects being masked by individual differences. In contrast, when we examined 136 non-significant effects from other domains, two novel non-parametric tests did reveal effects that were hidden by opposing individual results, though as we show, some of them might be driven by design-related factors. Taken together, five analysis approaches provide strong evidence for the restricted nature of unconscious processing effects not only across participants, but also across different trials within individuals. We provide analysis code and best-practice recommendations for testing for non-directional effects.

@@

Second, the *qualitative individual differences* approach (@rouder2021a; @haaf2019a; henceforth QUID) quantifies the relative support for the presence of “qualitative differences” in effects, that is, inter-individual differences in effect signs, by performing a Bayesian model comparison over a family of hierarchical models with different constraints [@haaf2019a]. Specifically, QUID models reaction times (RTs) using a standard linear model, where individual-level effects are treated as random effects. Then, to quantify the support for an effect while allowing for qualitative individual differences, it relies on Bayes Factors (BFs). The BF analysis compares evidence of a model that poses no constraints over the magnitude of individual-level effects, with a model in which all participants show no effect (see @rouder2021a for further model comparisons, aimed at examining different hypotheses regarding the character of population variability in effects).

@@

We applied the tests to our simulated data. For QUID, we used the default priors from the original publication [@rouder2021a]. For GNT and OANOVA, we used an $\alpha$ of `r oanova_alpha` to examine individual-level and group-level effects. For QUID, we considered $BF>`r bf_criteria`$ as evidence for an effect, $BF< \frac{1}{`r bf_criteria`}$ as evidence for no effect (*global null*), and values between these thresholds ($\frac{1}{`r bf_criteria`}\leq BF\leq `r bf_criteria`$) as inconclusive [@jeffreys1998a]. Reassuringly, all tests were able to differentiate between the two simulated scenarios, providing very strong evidence for an effect in the *non-directional differences* scenario, but not in the *global null* one. Specifically, the prevalence of effects on RT was clearly above the expected proportion of significant effects ($\alpha_{individual}$) in the *non-directional differences* simulation (using a two-sided t-test for the individual-level test; `r round(100 * nde_gnt_res$stat)`% significant effects, one-sided $CI_{95}$ = [`r round(100 * nde_gnt_res$ci_low)`, `r round(100 * nde_gnt_res$ci_high)`], p `r apa_p(nde_gnt_res$p)`), but this proportion was not higher than $\alpha_{individual}$ in the *global null* simulation (`r round(100 * sn_gnt_res$stat)`% significant effects, one-sided $CI_{95}$ = [`r round(100 * sn_gnt_res$ci_low)`, `r round(100 * sn_gnt_res$ci_high)`], p =`r apa_p(sn_gnt_res$p)`). Using the QUID method, a random effects model with individual-level effects was overwhelmingly preferred in the *non-directional differences* simulation ($BF$= `r formatC(nde_bf, format = 'e', 2)`), but a null model was preferred in the *global null* simulation ($BF$= `r round(sn_bf,2)`). Similarly, the OANOVA test revealed significant results in the *non-directional differences* scenario (F(`r nde_OANOVA_res$df_num`, `r nde_OANOVA_res$df_denom`) = `r round(nde_OANOVA_res$F,2)`, p `r paste0(ifelse(nde_OANOVA_res$p >= .001, '=', ''), apa_p(nde_OANOVA_res$p))`), and a non-significant effect in the *global null* simulation (F(`r sn_OANOVA_res$df_num`, `r sn_OANOVA_res$df_denom`) = `r round(sn_OANOVA_res$F,2)`, p `r paste0(ifelse(sn_OANOVA_res$p >= .001, '=', ''),apa_p(sn_OANOVA_res$p))`)^[Note that the unusually high degrees of freedom of OANOVA stem from modelling participants as fixed rather than random effects, as the global null model assumes no variability in effect sizes between participants.].

@@

Yet, the reviewed approaches also have some limitations that make it harder to draw firm conclusions based on their results. First, in contrast to frequentist tests within the Null Hypothesis Statistical Testing (NHST) tradition, QUID is designed to quantify relative evidence, and as such provides only weak^[Specifically, p($BF_{10}>x|H_0$) is never higher than $\frac{1}{x})$. Proof: assume that $BF_{10} > 3$ for observed values $v \in V$. That is, for every $v \in V$, $p(v|H_1)>3p(v|H_0)$. The probability of making a type-1 error $\alpha$ equals $p(v \in V|H_0)$. It follows that $p(v \in V|H_1)>3\alpha$. Since this probability cannot be higher than 1, $\alpha$ cannot exceed $\frac{1}{3}$ (for a similar derivation see the universal bound in @royall2000a] control over long-term error rates (the probability of finding a false positive result or missing a true result over an infinite number of tests, with the former being more critical to our point here). Such error control promises a much-needed ‘fool-proof’ method to infer the existence of unconscious processing effects without making too many mistakes in the long run [@lakens2020a; @kelter2021a].
